\input{preamble.tex}
\input{environments.tex}
\input{definitions.tex}

\externaldocument{../set-theory/set-theory}

\begin{document}

\title{Discrete Mathematics}
\author{Daniel Gonzalez Cedre}
\date{University of Notre Dame \\ Spring of 2023}
\maketitle

\setcounter{chapter}{3}
\chapter{Mathematical Induction}

\section{Weak Induction}

\begin{definition}[Well-Order]
    Let $X$ be a set with a relation $\leq$ defined on it.
    We say that $X$ is \emph{well-ordered} by this relation \ifandonlyif
    every non-empty subset of $X$ has a minimal element with respect to $\leq$.
    In other words, we say that $\leq$ is a \emph{well-order} on $X$
    \iffbydefn
    \[
        \pn*{\forall A \in \power{X}}\pn*{\exists a \in A}\pn*{\forall b \in A}{a \leq b}.
    \]
\end{definition}

\begin{theorem}[$\N$ is Well-Ordered]\label{thm:lep}
    \vspace{-\parskip}
    \[
        \pn*{\forall A \subseteq \N}\pn*{A \neq \emptyset \implies \pn*{\exists a \in A}\pn*{\forall b \in A}\pn*{a \leq b}}
    \]
    This says that every nonempty subset of $\N$ has a least element
    (according to the $\leq$ order defined on $\N$).
\end{theorem}
\begin{proof}
    This proof is left as an exercise.
\end{proof}

\begin{theorem}[Weak Induction]\label{thm:weakinduction}
    If $\varphi(\cdot)$ is a {\wff}, then
    \[
        \pn*{\forall n \in \N}\pn*{\varphi(n)}
        \iff \varphi(0) \meet \pn*{\forall k \in \N}\pn*{\varphi(k) \implies \varphi(k + 1)}.
    \]
\end{theorem}
\begin{proof}
    There are two fragments to this proof:
    the forward ($\implies$) direction and the backward ($\impliedby$) direction.
    \begin{case}[Fragment 1 ($\implies$)]
        Suppose that $\pn*{\forall n \in \N}\pn*{\varphi(n)}$.
        Since $0 \in \N$, we then obviously have $\varphi(0)$.
        Now, let $k \in \N$ and assume $\varphi(k)$.
        Since $k + 1 \in \N$, we know from our initial assumption that $\varphi(k + 1)$.
        Thus, we have $\pn*{\forall k \in \N}\pn*{\varphi(k) \implies \varphi(k + 1)}$,
        and we have reached both of our desired conclusions.
    \end{case}
    \begin{case}[Fragment 2 ($\impliedby$)]
        Assume $\varphi(0)$ and $\pn*{\forall k \in \N}\pn*{\varphi(k) \implies \varphi(k + 1)}$.
        Towards a contradiction,
        suppose that there is some $n \in \N$ such that $\neg \varphi(a)$.
        Consider $A \defn \set*{x \in \N \suchthat \neg \varphi(x)}$,
        which we clearly know exists by \autoref{ax:separation}.
        We know that $n \in A$ because we assumed that $\neg \varphi(n)$,
        which implies that $A \neq \emptyset$.
        Then, we can use \autoref{thm:lep} to conclude that there is a minimal element $a$ in $A$.

        Since we know that $\varphi(0)$, it follows that $a \neq 0$, so $a$ must be a successor number.
        This means there is a $b \in \N$ such that $b + 1 = a$.

        If $b \in A$, then that would mean that $a \leq b$ since $a$ is minimal in $A$.
        However, we know that $b < b + 1 = a$, so we would then have $a \leq b < a$. \contradiction
        Therefore, $b \not \in A$.
    \end{case}
    With these two directions proven, we finally have
    $\pn*{\forall n \in \N}\pn*{\varphi(n)}
    \iff \varphi(0) \meet \pn*{\forall k \in \N}\pn*{\varphi(k) \implies \varphi(k + 1)}$.
\end{proof}

Notice that the above theorem actually generalizes beyond just $\N$.
In fact, we can generalize the proof of \autoref{thm:weakinduction}
to \emph{any} well-ordered set $X$ by replacing $k + 1$
with the least element of the non-empty subset $X \setminus \set*{\ell \in X \suchthat \ell \leq k}$.

\newpage

Let's practice induction by proving the following few theorems.

\begin{theorem}[Gaussian Summation Formula]\label{thm:gaussiansum}
    $\displaystyle \pn*{\forall n \in \N}\pn*{\sum_{i = 0}^{n}i = \frac{n(n + 1)}{2}}$.
\end{theorem}
\begin{proof}
    We will prove the claim by induction on $n \in \N$.
    \begin{case}[Base Case]
        Observe that $\displaystyle \sum_{i = 0}^{0}i = 0 = \frac{0*(0 + 1)}{2}$.
        Therefore, the statement is satisfied at $0$.
    \end{case}
    \begin{case}[Inductive Step]
        Let $k \in \N$ and assume $\displaystyle \sum_{i = 0}^{k}i = \frac{k(k + 1)}{2}$
        (this is our \emph{inductive hypothesis}).
        Now, observe
        \begin{alignat*}{2}
            \sum_{i = 0}^{k + 1}i &= \pn*{\sum_{i = 0}^{k}i} + (k + 1) &&\text{ by definition}\\
                                  &= \pn*{\frac{k(k + 1)}{2}} + (k + 1)~~~~&&\text{ by the \emph{inductive hypothesis}}\\
                                  &= (k + 1)\pn*{\frac{k}{2} + 1} &&\text{ by the distributive property of multiplication}\footnotemark[1]\footnotemark[2]\\
                                  &= \frac{(k + 1)(k + 2)}{2} &&\text{ because } \frac{k}{2} + 1 = \frac{k}{2} + \frac{2}{2} = \frac{k + 2}{2}.\footnotemark[2]\footnotemark[3]
        \end{alignat*}
        Thus, we have that $\displaystyle \sum_{i = 0}^{k + 1}i = \frac{(k + 1)(k + 2)}{2}$, as desired.
    \end{case}
    Therefore, we can conclude that $\displaystyle \pn*{\forall n \in \N}\pn*{\sum_{i = 0}^{n}i = \frac{n(n + 1)}{2}}$.
\end{proof}

\begin{theorem}
    $3n \leq 3^n$ for all $n \in \N_+$.
\end{theorem}
\begin{proof}
    We will prove the claim by induction on $n \in \N_+$.
    \begin{case}[Base Case]
        Observe that $3 \cdot 0 = 0 \leq 1 = 3^0$.
    \end{case}
    \begin{case}[Inductive Step]
        Let $k \in \N_+$ and assume $3 \cdot k \leq 3^k$
        (this is our \emph{inductive hypothesis}).
        Now, observe
        \begin{alignat*}{2}
            3 \cdot (k + 1) &= 3 \cdot k + 3 &&\text{ by the distributive property of multiplication} \\
                            &\leq 3^k + 3 &&\text{ by the \emph{inductive hypothesis}} \\
                            &\leq 3^k + 3^k &&\text{ because $1 \leq k \implies a^1 \leq a^k$ if $a < 0$}\footnotemark[1]\footnotemark[2] \\
                            &\leq 3^k + 3^k + 3^k~~~~&&\text{ since $a^k$ is always positive if $0 < a$}\footnotemark[1]\footnotemark[2] \\
                            &= 3 \cdot 3^k &&\text{ because $\textstyle \sum_{i = 1}^{m}a = m \cdot a$ for any $a$}\footnotemark[1]\footnotemark[2] \\
                            &= 3^{k + 1} &&\text{ because $a^ba^c = a^{b + c}$ for any $a$}.\footnotemark[1]\footnotemark[2]
        \end{alignat*}
        So, $3 \cdot (k + 1) \leq 3^{k + 1}$, as desired.
    \end{case}
    Therefore, $\pn*{\forall n \in \N}\pn*{3n \leq 3^n}$.
\end{proof}

\footnotetext[1]{These ``basic grade-school'' algebraic properties will now be assumed without special mention.}
\footnotetext[2]{This is true in any \emph{ordered semiring} where exponentiation is defined in terms of multiplication.}
\footnotetext[3]{This is true in any \emph{field}.}

\begin{theorem}
    $\displaystyle \sum_{i = 0}^{n}2^i = 2^{n + 1} - 1$ for all $n \in \N$.
\end{theorem}
\begin{proof}
    We will prove the claim by induction on $n \in \N_+$.
    \begin{case}[Base Case]
        Observe that $\displaystyle \sum_{i = 0}^{0}2^i = 2^0 = 1 = 2 - 1 = 2^{1 + 1} - 1$
    \end{case}
    \begin{case}[Inductive Step]
        Let $k \in \N_+$ and assume that $\displaystyle \sum_{i = 0}^{k}2^i = 2^{k + 1} - 1$
        (this is our \emph{inductive hypothesis}).
        Observe that
        \begin{alignat*}{2}
            \sum_{i = 0}^{k + 1}2^i &= \pn*{\sum_{i = 0}^{k}2^i} + 2^{k + 1} &&\text{ by definition} \\
                                    &= \pn*{2^{k + 1} - 1} + 2^{k + 1}~~~~&&\text{ by the \emph{inductive hypothesis}} \\
                                    &= 2 \cdot 2^{k + 1} - 1 &&\text{ using some basic algebra} \\
                                    &= 2^{k + 2} - 1 &&\text{ using some basic algebra}.
        \end{alignat*}
        So, we get $\displaystyle \sum_{i = 0}^{k + 1}2^i = 2^{k + 1} - 1$, as desired.
    \end{case}
    Therefore, we can conclude $\displaystyle \pn*{\forall n \in \N}\pn*{\sum_{i = 0}^{n}2^i = 2^{n + 1} - 1}$.
\end{proof}

\end{document}
