\input{preamble.tex}

\begin{document}
\begin{center}
    \textsc{\huge Problem Set 1}\\
    \textsc{\large Discrete Mathematics}\\
    {\color{red!75!black}Due: $27$\textsuperscript{th} of January, $2023$}
\end{center}

\begin{enumerate}
    \item
        Which of the following are propositions and which are not? Justify your answer with an argument.
        \begin{enumerate}
            \item
                Santa Claus is real and he lives in the South Pole.
            \item
                This sentence is true.
            \item
                If this sentence is false, then 7 is a prime number.
            \item
                This sentence is false.
            \item
                Every proposition is either true or false.
            \item
                If this sentence is true, then $12$ is a prime number.
            \item
                The capitial of Spain is Madrid.
            \item
                There is a set $x$ that contains itself as an element.
            \item
                The set of all sets that do not contain themselves contains itself.
            \item
                $\sqrt{2}$ is a morally complicated number.
        \end{enumerate}
    \item \label{prob:2}
        For \hyperref[prob:2]{problem~2} and \hyperref[prob:3]{problem~3},
        consider the following definitions:
        \begin{align*}
            N &\defn \text{``There is an alien aboard the Nostromo.''}\\
            E &\defn \text{``The crewmembers go on an expedition.''}\\
            A &\defn \text{``A crewmember was infected by an alien.''}\\
            R &\defn \text{``The crewmembers return from their expedition.''}\\
            S &\defn \text{``The crewmembers survive.''}\\
            K &\defn \text{``Ripley kills an alien.''}
        \end{align*}

        Translate the following English-language sentences into the propositional logic:
        \begin{enumerate}
            \item
                If there is no alien aboard the Nostromo,
                then Ripley will not kill an alien.
            \item
                The crewmembers went on an expedition but did not return.
            \item
                Unless Ripley kills the alien, the crewmembers won't survive.
            \item
                A crewmember was infected by an alien
                and then the crewmembers returned from their expedition.
            \item
                Although there is an alien aboard the Nostromo,
                the crewmembers will survive.
            \item
                It is sufficient that Ripley kills an alien
                for the crewmembers to survive.
            \item
                It is necessary that Ripley kills an alien
                for the crewmembers to survive.
            \item
                Whether or not Ripley kills an alien,
                the crewmembers will not survive.
            \item
                It is not the case that
                there is an alien aboard the Nostromo
                or Ripley kills an alien.
            \item
                The crewmembers went on an expedition
                -- wherein a crewmember was infected by an alien --
                and then returned,
                so an alien is aboard the Nostromo;
                yet, the crewmembers will survive.
        \end{enumerate}

    \item \label{prob:3}
        Translate the following sentences of the propositional logic into English:
        \begin{enumerate}
            \item
                $\neg K \rightarrow \neg S$
            \item
                $\neg (S \rightarrow K)$
            \item
                $(N \meet K) \rightarrow S$
            \item
                $(R \join \neg R) \meet \neg (R \meet \neg R)$
            \item
                $R \leftrightarrow N$
        \end{enumerate}
\end{enumerate}
\end{document}
